---
title: "Untitled"
output: html_document
date: "2025-04-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Cleaning up the data
```{r}
#Reading in the txt file and coverting it to a dataframe
lines <- readLines("~/Downloads/s_MTBLS5568.txt")
split_line <- strsplit(lines, "\t")
samples <- do.call(rbind, lapply(split_line, function(x) as.data.frame(t(x), stringAsFactor=FALSE)))
# I want the fist row of the coloumn to be the columnames
colnames(samples) <- samples[1, ]
samples <- samples[-1, ]
#Renaming columns so they are easier to work with
colnames(samples) <- make.unique(colnames(samples))
samples <- samples%>%
  rename(source=`Source Name`, sample_type=`Characteristics[Sample type]`, sample_name=`Sample Name`, ethnicity=`Factor Value[Ethnic group]`, WPI=`Factor Value[WPI]`, time=`Factor Value[Time]`, participant_id=`Comment[Participant ID]`, age=`Factor Value[Age]`, BMI=`Factor Value[BMI]`, MRI_pf=`Factor Value[MRI pancreas fat]`)
plot(samples$MRI_pf~samples$BMI)
plot(samples$BMI~samples$age)
plot(samples$MRI_pf~samples$age)
```
```{r}
group_by(samples$ethnicity)
plot(samples$ethnicity~samples$MRI_pf)
plot(samples$ethnicity~samples$BMI)
```

## metabolite data 

```{r}
metabolite_data <- read.table("~/Downloads/m_MTBLS5568_LC-MS_neg_hilic_metabolite_profiling_v2_maf.tsv", header = TRUE, sep="\t")
```

```{r}
metabolites_l <- metabolite_data%>%
  pivot_longer(cols=starts_with("X"), names_to="source", values_to = "value")%>%
  pivot_wider(names_from= metabolite_identification, values_from = value)
```

New data frame with selected metabolites: 
```{r}
metabolites_short <- metabolites_l%>%
  select(source, `L-Threonine`, `Citric acid`, `Uric acid`, `L-Cystine`, `L-Tyrosine`, `L-Histidine`)
```
Remove X from source name: 
```{r}
metabolites_short <- data.frame(lapply(metabolites_short, function(x) gsub("^X", "", x)))
```

Now I can merge the two dataframes: 
```{r}
merge_df <- merge(samples, metabolites_short, by="source", all.x=TRUE)
```

Now I can use merged_df, when looking at metabolites and ethnicity and other variables. 
Plot time against citric.acid and group by WPI:
## Why is this not working???

```{r}
clean_data <- merge_df%>%
  filter(!is.na(Citric.acid))

mean_data <- clean_data%>%
  group_by(time, WPI)%>%
  summarise(mean_citric_acid=mean(Citric.acid, na.rm=TRUE))

print(ggplot(mean_data, aes(x=time, y=mean_citric_acid, color=WPI))+
  geom_point()+
  geom_line()+
  labs(title="Citric Acid", 
       c= "Time", 
       y="Citric Acid"))
```
Might want to group by both WPI and MRI - how do I do that?
and classify MRI as high and low 

Also it could be interesting to see how BMI and WPI grouping changes in compariset to MRI
and then looking a ethnic differences (first see if there are wthnic differences in MRI and BMI) but then also if there are differences in metablome. 

Maybe end goal is to find out what affects the metablome (potential differences) - is it BMI, ethnicity, MRI (age)
In the study they say that ethnicity was a stronger predictor for metablome than MRI
What is the simples model (least variables) that can explain correlation between metablom(conc of metabolites) and WPI intake? so that would be an anova with metablome as a response variable and WPI intake as the groups. and then find out if we need subgroups

Questions?: 
- what does the feature exclusion pipeline mean for the data reliability? 

Få overblik over source - alle kan være b1 eller b2.  